Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-16 write-ups/PLDI/paper.tex
Fix hyphenation

@@

\item The model is backwards-compatible: if $\offerkill$/$\offerfork$ are ignored, a cooperative kernel will behave
  exactly as a regular kernel does on current GPUs.

\item Cooperative kernels can be implemented over the occupancy-bound
  execution model that current GPUs provide: our prototype implementation requires
  no additional hardware or driver support.  Thus the model should be easy for vendors to implement directly.

@@

\myfiglong\ref{fig:workstealing} shows a simplified version of a work
stealing kernel. Each thread receives a pointer to the
task queues, in global memory, initialized by the
host to contain the initial tasks. Each thread identifies
its workgroup via $\getgroupid()$
(line~\ref{line:wksteal:getgroupid}), which is used as a queue id to access the relevant task queue. The
$\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
is called by all threads of a workgroup.  Inside
$\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
the \emph{master thread}, tries to pop a
task from the workgroup's queue, or to steal a task from a different
queue. The master thread uses local memory inside
$\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
all threads of its workgroup. If a task was obtained, then all the
workgroup threads participate in the task processing (line~\ref{line:wksteal:processtask}).
This may lead to the creation of further tasks, which
the master thread pushes to the workgroup's queue.

@@

The application iterates as long as the current frontier contains
nodes to process (line~\ref{line:graph:iterate}). At each frontier,
the nodes to be processed are evenly distributed between
threads through \emph{stride} based processing. 
%
In this case, the stride is the total number of threads, obtained via 
$\getglobalsize()$.  Each thread processes nodes with the following sequence of indices:
\code{$\keyword{tid}$+$\keyword{stride}$*0},
\code{$\keyword{tid}$+$\keyword{stride}$*1},
\code{$\keyword{tid}$+$\keyword{stride}$*2}, etc., where $\keyword{tid}$ is a thread's global id.
A thread processes a node via $\keyword{process\_node}$, which may (a) push nodes to process in the next frontier to
$\keyword{out\_nodes}$ and (b) use the current frontier, $\keyword{level}$, in
the computation. After processing the frontier, the threads swap their
node array pointers (line~\ref{line:graph:swap}).

@@

When executed by a cooperative
kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
active workgroups.  This is in contrast to $\getnumgroups()$ for
regular kernels, which returns the fixed number of workgroups that
were requested at kernel launch time (see \mysec\ref{sec:opencl}).

@@

Our $\globalbarrier$ primitive is a kernel-level function: if it
appears in conditional code then it must be reached by \emph{all}
threads executing the cooperative kernel.  On reaching a
$\globalbarrier()$, a thread waits until all threads have arrived at
the barrier.  Once all threads have arrived, the threads may proceed
past the barrier with the guarantee that all global memory accesses
issued before the barrier have completed.  The $\globalbarrier$
primitive can be implemented by adapting an inter-workgroup barrier
design, e.g.~\cite{XF10}, to take account of a growing and shrinking number of workgroups, and the atomic operations provided by
the OpenCL 2.0 memory model enable a memory-safe
implementation~\cite{DBLP:conf/oopsla/SorensenDBGR16}.  However, implementing such a barrier is
involved, hence why we include this function as a primitive.

@@

\paragraph{A changing workgroup count}  Unlike in regular OpenCL,
the value returned by $\getnumgroups()$ is not fixed during the
lifetime of a cooperative kernel: it corresponds to the active group count $M$, which changes as workgroups execute
$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  The value returned by $\getglobalsize()$
is similarly subject to change.
A cooperative kernel must thus be written in a manner that is
robust to changes in the values returned by these functions.

In general, their volatility means that use of these functions should be avoided.
However, the situation is more stable if a cooperative kernel does not call
$\offerkill$ and $\offerfork$ directly, so that only $\resizingglobalbarrier$ can affect the number of active
workgroups.  Then, at any point
during execution, the threads of a kernel are executing between some
pair of resizing barrier calls, which we call a \emph{resizing barrier
  interval} (considering the kernel entry and exit points
conceptually to be special cases of resizing barriers).  The active workgroup count is constant within each resizing
barrier interval, so that 
$\getnumgroups()$ and $\getglobalsize()$ return stable values during such intervals.
%
As we illustrate below for graph traversal, this can be exploited by algorithms that perform strided
data processing.

@@

\paragraph{Adapting graph traversal} 
\myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the
graph traversal kernel of \myfig\ref{fig:graphsearch} from
\mysec\ref{sec:openclexamples}.  On lines~\ref{line:cgraph:resizing1}
and ~\ref{line:cgraph:resizing2}, we change the original global
barriers into a resizing barriers. Several variables are marked to be
transmitted in the case of workgroups joining at the resizing barriers
(lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
\ref{line:cgraph:transmit3}): $\keyword{level}$ must be restored so
that new workgroups know which frontier they are processing;
$\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ must be restored so
that new workgroups know which of the node arrays to use for input and
output. Lastly, the static work distribution of the original kernel is
no longer valid in a cooperative kernel. This is because the stride
(which is based on $M$) may change after each resizing barrier
call. To fix this, we re-distribute the work after each resizing
barrier call by recomputing the thread id and stride
(lines~\ref{line:cgraph:rechunking1} and
\ref{line:cgraph:rechunking2}). This example exploits the fact that
the cooperative kernel does not issue $\offerkill$ nor $\offerfork$
directly: the value of $\keyword{stride}$ obtained from
$\getglobalsize()$ at line~\ref{line:cgraph:rechunking2} is stable
until the next resizing barrier at line~\ref{line:cgraph:resizing1},
and in particular can be safely used inside the $\keyword{for}$ loop.

@@

We can greatly reduce the gather time by providing a new $\keyword{query}()$
function for a workgroup executing a cooperative kernel, which returns the number of workgroups that the scheduler needs to obtained
from the cooperative kernel.  
%
The programmer cannot call $\keyword{query}$ directly;
it is only invoked from within the resizing barrier implementation.
%
A resizing barrier can now be implemented
as follows: (1) the master waits for all slaves to arrive; (2) the master calls $\offerfork$ and
commands the new workgroups to be slaves; (3) the master calls $\keyword{query}$, obtaining a value $W$;
(4) the master releases the slaves, broadcasting the value $W$ to them; (5) workgroups with ids larger than $M-W$ spin, calling 
$\offerkill$ repeatedly until the scheduler claims them---we know from $\keyword{query}$ that the scheduler will eventually do so.
We show in \mysec\ref{sec:responsiveness} that the barrier using
query greatly reduces the gather time in practice.

@@

The $\textsc{Barrier}$ rule requires that \emph{every} thread
executing the kernel reaches $\globalbarrier()$ an identical following
statement.  This reflects that fact that $\globalbarrier()$ is a
kernel-level function.  Each thread then skips over the
$\globalbarrier()$ call, and the $\sync$ function is applied to yield
a set of kernel states that can arise due to memory synchronization
taking place.  An arbitary member of this set is selected as the next
kernel state.

@@

We opted for a semantics whereby only the active workgroup with the
highest id can be killed.  This has an appealing property: it means
that the ids of active workgroups are contiguous, which is important
for processing of contiguous data.  The cooperative graph traversal
algorithm of \myfig\ref{fig:cgraphsearch} illustrates this: the
algorithm is prepared for $\getglobalsize()$ to change after each
resizing barrier call, but depends on the fact that $\getglobalid()$
returns a contiguous range of thread ids.

>>>>

\usepackage[shortcuts]{extdash} % hyphenation, see http://stackoverflow.com/questions/2193307/how-do-i-get-latex-to-hyphenate-a-word-that-contains-a-dash#8649145

@@

\item The model is backwards-compatible: if $\offerkill$ and $\offerfork$ are ignored, a cooperative kernel will behave
  exactly as a regular kernel does on current GPUs.

\item Cooperative kernels can be implemented over the occupancy\-/bound
  execution model that current GPUs provide: our prototype implementation requires
  no additional hardware or driver support.  Thus the model should be easy for vendors to implement directly.

@@

\myfiglong\ref{fig:workstealing} shows a simplified version of a work
stealing kernel. Each thread receives a pointer to the
task queues, in global memory, initialized by the
host to contain the initial tasks. Each thread identifies
its workgroup via $\getgroupid$
(line~\ref{line:wksteal:getgroupid}), which is used as a queue id to access the relevant task queue. The
$\mathsf{pop\_or\_steal}$ function (line~\ref{line:wksteal:poporsteal})
is called by all threads of a workgroup.  Inside
$\mathsf{pop\_or\_steal}$ only the thread with a local id equal to $0$,
the \emph{master thread}, tries to pop a
task from the workgroup's queue, or to steal a task from a different
queue. The master thread uses local memory inside
$\mathsf{pop\_or\_steal}$ to ensure that the returned value is valid for
all threads of its workgroup. If a task was obtained, then all the
workgroup threads participate in the task processing (line~\ref{line:wksteal:processtask}).
This may lead to the creation of further tasks, which
the master thread pushes to the workgroup's queue.

@@

The application iterates as long as the current frontier contains
nodes to process (line~\ref{line:graph:iterate}). At each frontier,
the nodes to be processed are evenly distributed between
threads through \emph{stride} based processing. 
%
In this case, the stride is the total number of threads, obtained via 
$\getglobalsize$.  Each thread processes nodes with the following sequence of indices:
\code{$\keyword{tid}$+$\keyword{stride}$*0},
\code{$\keyword{tid}$+$\keyword{stride}$*1},
\code{$\keyword{tid}$+$\keyword{stride}$*2}, etc., where $\keyword{tid}$ is a thread's global id.
A thread processes a node via $\keyword{process\_node}$, which may (a) push nodes to process in the next frontier to
$\keyword{out\_nodes}$ and (b) use the current frontier, $\keyword{level}$, in
the computation. After processing the frontier, the threads swap their
node array pointers (line~\ref{line:graph:swap}).

@@

When executed by a cooperative
kernel, $\getnumgroups$ returns $M$, the \emph{current} number of
active workgroups.  This is in contrast to $\getnumgroups$ for
regular kernels, which returns the fixed number of workgroups that
were requested at kernel launch time (see \mysec\ref{sec:opencl}).

@@

Our $\globalbarrier$ primitive is a kernel-level function: if it
appears in conditional code then it must be reached by \emph{all}
threads executing the cooperative kernel.  On reaching a
$\globalbarrier$, a thread waits until all threads have arrived at
the barrier.  Once all threads have arrived, the threads may proceed
past the barrier with the guarantee that all global memory accesses
issued before the barrier have completed.  The $\globalbarrier$
primitive can be implemented by adapting an inter-workgroup barrier
design, e.g.~\cite{XF10}, to take account of a growing and shrinking number of workgroups, and the atomic operations provided by
the OpenCL 2.0 memory model enable a memory-safe
implementation~\cite{DBLP:conf/oopsla/SorensenDBGR16}.  However, implementing such a barrier is
involved, hence why we include this function as a primitive.

@@

\paragraph{A changing workgroup count}  Unlike in regular OpenCL,
the value returned by $\getnumgroups$ is not fixed during the
lifetime of a cooperative kernel: it corresponds to the active group count $M$, which changes as workgroups execute
$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  The value returned by $\getglobalsize$
is similarly subject to change.
A cooperative kernel must thus be written in a manner that is
robust to changes in the values returned by these functions.

In general, their volatility means that use of these functions should be avoided.
However, the situation is more stable if a cooperative kernel does not call
$\offerkill$ and $\offerfork$ directly, so that only $\resizingglobalbarrier$ can affect the number of active
workgroups.  Then, at any point
during execution, the threads of a kernel are executing between some
pair of resizing barrier calls, which we call a \emph{resizing barrier
  interval} (considering the kernel entry and exit points
conceptually to be special cases of resizing barriers).  The active workgroup count is constant within each resizing
barrier interval, so that 
$\getnumgroups$ and $\getglobalsize$ return stable values during such intervals.
%
As we illustrate below for graph traversal, this can be exploited by algorithms that perform strided
data processing.

@@

\paragraph{Adapting graph traversal} 
\myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the
graph traversal kernel of \myfig\ref{fig:graphsearch} from
\mysec\ref{sec:openclexamples}.  On lines~\ref{line:cgraph:resizing1}
and ~\ref{line:cgraph:resizing2}, we change the original global
barriers into a resizing barriers. Several variables are marked to be
transmitted in the case of workgroups joining at the resizing barriers
(lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
\ref{line:cgraph:transmit3}): $\keyword{level}$ must be restored so
that new workgroups know which frontier they are processing;
$\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ must be restored so
that new workgroups know which of the node arrays to use for input and
output. Lastly, the static work distribution of the original kernel is
no longer valid in a cooperative kernel. This is because the stride
(which is based on $M$) may change after each resizing barrier
call. To fix this, we re-distribute the work after each resizing
barrier call by recomputing the thread id and stride
(lines~\ref{line:cgraph:rechunking1} and
\ref{line:cgraph:rechunking2}). This example exploits the fact that
the cooperative kernel does not issue $\offerkill$ nor $\offerfork$
directly: the value of $\keyword{stride}$ obtained from
$\getglobalsize$ at line~\ref{line:cgraph:rechunking2} is stable
until the next resizing barrier at line~\ref{line:cgraph:resizing1},
and in particular can be safely used inside the $\keyword{for}$ loop.

@@

We can greatly reduce the gather time by providing a new $\keyword{query}$
function for a workgroup executing a cooperative kernel, which returns the number of workgroups that the scheduler needs to obtained
from the cooperative kernel.  
%
The programmer cannot call $\keyword{query}$ directly;
it is only invoked from within the resizing barrier implementation.
%
A resizing barrier can now be implemented
as follows: (1) the master waits for all slaves to arrive; (2) the master calls $\offerfork$ and
commands the new workgroups to be slaves; (3) the master calls $\keyword{query}$, obtaining a value $W$;
(4) the master releases the slaves, broadcasting the value $W$ to them; (5) workgroups with ids larger than $M-W$ spin, calling 
$\offerkill$ repeatedly until the scheduler claims them---we know from $\keyword{query}$ that the scheduler will eventually do so.
We show in \mysec\ref{sec:responsiveness} that the barrier using
query greatly reduces the gather time in practice.

@@

The $\textsc{Barrier}$ rule requires that \emph{every} thread
executing the kernel reaches $\globalbarrier$ an identical following
statement.  This reflects that fact that $\globalbarrier$ is a
kernel-level function.  Each thread then skips over the
$\globalbarrier$ call, and the $\sync$ function is applied to yield
a set of kernel states that can arise due to memory synchronization
taking place.  An arbitary member of this set is selected as the next
kernel state.

@@

We opted for a semantics whereby only the active workgroup with the
highest id can be killed.  This has an appealing property: it means
that the ids of active workgroups are contiguous, which is important
for processing of contiguous data.  The cooperative graph traversal
algorithm of \myfig\ref{fig:cgraphsearch} illustrates this: the
algorithm is prepared for $\getglobalsize$ to change after each
resizing barrier call, but depends on the fact that $\getglobalid$
returns a contiguous range of thread ids.
